Best Practices for Alchemical Free Energy Calculations [Article v \versionnumber]

### 2020-11-03 paper/manuscript.tex
adding version 1.0

\newcommand{\versionnumber}{1.0} % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/michellab/alchemical-best-practices}} %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Alchemical Free Energy Calculations [Article v \versionnumber]}
\author[1*]{Antonia S. J. S. Mey}
\author[7]{Bryce K. Allen}
\author[2]{Hannah E. Bruce Macdonald}
\author[2*]{John D. Chodera}
\author[1,10]{Maximilian Kuhn}
\author[1]{Julien Michel}
\author[3*]{David L. Mobley}
\author[11]{Levi N. Naden}
\author[4]{Samarjeet Prasad}
\author[2,8]{Andrea Rizzi}
\author[1]{Jenke Scheen}
\author[6*]{Michael R. Shirts}
\author[9]{Gary Tresadern}
\author[7]{Huafeng Xu}
%
\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[2]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[3]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, USA}
\affil[4]{National Institutes of Health, Bethesda, MD, USA}
\affil[6]{University of Colorado Boulder, Boulder, CO, USA}
\affil[7]{Silicon Therapeutics, Boston, MA, USA}
\affil[8]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340,Belgium}
\affil[10]{Cresset, Cambridgeshire, UK}
\affil[11]{Molecular Sciences Software Institute, Blacksburg VA, USA}
%
\corr{antonia.mey@ed.ac.uk}{ASJSM}
\corr{john.chodera@choderalab.org}{JDC}
\corr{dmobley@mobleylab.org}{DLM}
\corr{michael.shirts@colorado.edu}{MRS}

\orcid{Antonia S. J. S. Mey}{0000-0001-7512-5252}
\orcid{Bryce Allen}{0000-0002-0804-8127}
\orcid{Hannah E. Bruce Macdonald}{0000-0002-5562-6866}
\orcid{John D. Chodera}{0000-0003-0542-119X}
\orcid{Maximilian Kuhn}{0000-0002-2811-3934}
\orcid{Julien Michel}{0000-0003-0360-1760}
\orcid{David L. Mobley}{0000-0002-1083-5533}
\orcid{Levi N. Naden}{0000-0002-3692-5027}
\orcid{Samarjeet Prasad}{0000-0001-8320-6482}
\orcid{Andrea Rizzi}{0000-0001-7693-2013}
\orcid{Jenke Scheen}{0000-0001-9781-0445}
\orcid{Michael R. Shirts}{0000-0003-3249-1097}
\orcid{Gary Tresadern}{0000-0002-4801-1644}
\orcid{Huafeng Xu}{0000-0001-5447-0452}

\blurb{This LiveCoMS document is maintained online on GitHub at \githubrepository; to provide feedback, suggestions, or help improve it, please visit the GitHub repository and participate via the issue tracker.}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PUBLICATION INFORMATION
%%% Fill out these parameters when available
%%% These are used when the "pubversion" option is invoked
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\pubDOI{10.XXXX/YYYYYYY}
\pubvolume{<volume>}
\pubyear{<year>}
\articlenum{<number>}
\datereceived{Month, Day, Year}
\dateaccepted{Month, Day, Year}

@@

In this paper, we review current best practices for several popular application domains of alchemical free energy calculations, including relative and absolute small molecule binding free energy calculations to biomolecular targets.
%\todo[inline, color=green!20]{JDC: Should we migrate this document to \url{https://github.com/alchemistry}? ASJSM: Let's migrate with the submitted version, but leave it here for now?}
\end{abstract}
\end{frontmatter}

@@

Subsequent work in the 2000s led to improved implementations of alchemical methods in popular biomolecular simulation packages~\cite{shirts2003extremely,shirts2005solvation,vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}. 
 This foundational work, combined with the methodological, technological, and hardware improvements of the last 5--10 years, has led to an explosion of interest and direct commercial application of these technologies~\cite{wang2015accurate, fratev2019improved, schindler2020largescale, cournia2017relative, sherborne2016collaborating, kuhn2020assessment}.

@@

As some of the theoretical background can seem daunting, we do, however, provide a guide to the essential theory behind alchemical free energy calculations in Sec.~\ref{sec:theory}.
In the remainder of this paper, we will cover topics that are key to the preparation( Sec.~\ref{sec:prerequisites}), choice and use of correct protocols (Sec.~\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used in the analysis of alchemical calculations (Sec.~\ref{sec:data_analysis}). 
Particular focus will be given to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative).

While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future Best Practices documents focusing on specific issues. 
Below, we provide a non-exhaustive list of topics we have \emph{not} addressed, along with some references to provide starting points on these more advanced topics:
\begin{itemize}
\item covalent inhibition~\cite{lameira2019predicting}
\item free energies of mutation of protein side chains~\cite{gapsys2016accurate, aldeghi2018accurate}
\item nonspecific binding or multiple binding sites~\cite{gill2018binding}
\item approximate and often less accurate endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
\item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
\item forcefield dependence for protein, ligand, ions, co-solvents, and co-factors. A number of different studies have looked at the influence of force fields and it is assumed the user has made an adequate choice for the system under study~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}. 
\end{itemize}

For convenience we have also compiled a list of common acronyms and common symbols used throughout this paper.
\begin{tcolorbox}[title=Acronyms, colback=blue!10!white]
    {\bf CPU} --- Central Processing Unit\\
     {\bf BAR} --- Bennett Acceptance Ratio\\
     {\bf FEP} --- Free Energy Perturbation\\
     {\bf GPCR} --- G-Protein Coupled Receptor\\
     {\bf GPU} --- Graphics Processing Unit\\
     {\bf MBAR} --- Multistate Bennett Acceptance Ratio\\
     {\bf MCSS} --- Maximum Common Substructure\\
     {\bf MD} --- Molecular Dynamics\\
     {\bf RMSE} --- Root Mean Square Error\\
     {\bf MUE} --- Mean unsigned error\\
     {\bf SAR} --- Structure-Activity Relationships\\
     {\bf TI} --- Thermodynamic Integration
\end{tcolorbox}

\begin{tcolorbox}[title=List of Symbols, colback=green!10!white]
$L$ and $R$ --- generic names for ligand and receptor\\
$K_b^{\circ}$ --- binding  constant \\
$c^{\circ}$ --- standard state concentration \\
$U$ --- potential energy\\
$u$ --- reduced potential describing a thermodynamic state \\
$\Delta G$ --- Gibbs free energy (isothermal isobaric ensemble)\\
$\Delta A$ --- Helmholtz free energy (canonical ensemble)\\
$\Delta f$ --- reduced (dimensionless) free energy \\
$\Delta \hat{f}$ --- estimate from an estimator for the reduced free energy\\
$\Gamma$ --- conformation space accessible by simulations \\
$\vec{q}$ --- vector of a single configuration, i.e. $x$, $y$, $z$ coordinates of the simulation system\\
$k_B$ --- Boltzmann constant \\
$Z$ --- partition function \\
$p$ --- pressure \\
$\mu$ --- chemical potential (grand canonical ensemble)\\
$T$ --- temperature \\
$\beta \equiv (k_B T)^{-1}$ --- inverse thermal energy \\
$\vec{\lambda}$ --- alchemical progress parameter, which may be multidimensional \\
$g$ --- statistical inefficiency\\
$\mathcal{O}$ --- overlap matrix\\
$C_t$ --- discrete-time-normalized fluctuation auto-correlation function\\
$\tau _{eq}$ --- integrated auto-correlation time\\
$t_0$ --- equilibration time
\end{tcolorbox}

@@

\subsection{Simulating binding events of receptor-drug systems can be computationally expensive}
Suppose you want to compute the binding affinity, or free energy of binding, of a ligand $L$ to a receptor $R$, given by:
\begin{equation}
R+L \leftrightharpoons RL.
\end{equation}
The binding constant ($K_b^{\circ}$) is given by the law of mass action as the ratio of concentrations of product $[RL]$ and reactants $[R]$, $[L]$:
\begin{equation}
 K_b^{\circ} = c^{\circ}\frac{[RL]}{[L][R]}.
\end{equation}
The standard state concentration $c^\circ$ depends on the reference state, but it is usually set to $1\,\mathrm{mol}/\mathrm{L}$ assuming a constant pressure of $1\,\mathrm{atm}$ (see also Sec.~\ref{sec:standardstate-restraints}).
Thus, the Gibbs free energy of binding $\Delta G_{\mathrm{bind}}$ is given by:
\begin{equation}
    \Delta G_{\mathrm{bind},L} = -k_BT\ln K_b^{\circ},
\end{equation}
where $k_B$ is the Boltzmann constant and $T$ the temperature of the system.

\paragraph{The free energy of binding can be expressed as a ratio of partition functions}
A natural, though generally very computationally expensive, way to estimate the equilibrium constant is by directly simulating several binding and unbinding events and computing the probability of finding the receptor-ligand system in the bound state, $P(RL)$, or the unbound state, $P(R+L)$.
Assuming the volume change upon binding to be negligible, which is often the case at $1~\mathrm{atm}$ due to the incompressibility of water, then the Gibbs free energy $\Delta G_{\mathrm{bind}, L}$ is approximately equal to the Helmholtz free energy $\Delta A_{\mathrm{bind}, L}$, and we can simulate the system in a box of volume $V$ to obtain~\cite{jong2011determining}
\begin{equation}\label{eq:binding-free-energy-from-bound-unbound-probability-ratio}
    \Delta G_{\mathrm{bind},L} \approx \Delta A_{\mathrm{bind}, L} = - k_B T  \left( \mathrm{ln} \frac{P(RL)}{P(R+L)} + \ln \left( c^\circ N_{\mathrm{Av}} V \right) \right) \, ,
\end{equation}
where $N_{\mathrm{Av}}$ is the Avogadro number, and the last term corrects for the simulated concentration being different than the standard concentration.
Let $\Gamma_{\mathrm{bound}}$ and $\Gamma_{\mathrm{unbound}}$ be the set of receptor-ligand conformations $\vec{q}$ that we consider bound and unbound respectively.
The probability of a conformation $\vec{q}$ is given by the Boltzmann probability density function
\begin{equation}
    P(\vec{q}) = \frac{\exp\left( \beta U(\vec{q}) \right)}{\int_{\Gamma} \exp\left( \beta U(\vec{q}) \right) \, d\vec{q}} \, ,
    \label{eq:conf_probability}
\end{equation}
where $\beta = (k_B T)^{-1}$ is the inverse temperature, $U(\vec{q})$ is the potential energy of conformation $\vec{q}$, and the integration is over the set of all possible conformations accessible in the simulation box volume $\Gamma$, with $\Gamma_{\mathrm{bound}},\Gamma_{\mathrm{unbound}} \subset \Gamma$.
If the simulation is long enough, we expect the fraction of conformations $\vec{q}$ found in the bound state to converge to
\begin{equation}
    P(RL) = \int_{\Gamma_{\mathrm{bound}}} P(\vec{q}) \, d\vec{q} = \frac{\int_{\Gamma_{\mathrm{bound}}} \exp\left( \beta U(\vec{q}) \right) \, d\vec{q}}{\int_{\Gamma} \exp\left( \beta U(\vec{q}) \right) \, d\vec{q}} \, .
\end{equation}
After similar considerations for $P(R+L)$, we find that the ratio of visited bound and unbound conformations, in the limit of long simulations, should converge to
\begin{equation}\label{eq:bound-unbound-probability-ratio}
    \frac{P(RL)}{P(R+L)} = \frac{\int_{\Gamma_{\mathrm{bound}}} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}}{\int_{\Gamma_{\mathrm{unbound}}} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}} = \frac{Z(RL)}{Z(R+L)} \, ,
\end{equation}
where we have defined the \textit{configurational integral} or \textit{configurational partition function} as $Z(\mathrm{state}) \equiv \int_{\Gamma_{\mathrm{state}}} \exp\left(-\beta U(\vec{q})\right) \, d\vec{q}$.

\paragraph{Simulating binding events is computationally expensive}
While simulating binding events has been used to estimate binding affinities~\cite{jong2011determining,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo2016adaptive,votapka2017seekr,doerr2014onthefly,plattner2015protein,dixon2018predicting}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the microsecond timescale even for millimolar binders~\cite{pan2017quantitative} and reaches the micosecond to second timescale for a typical drug~\cite{basavapathruni2012conformational,hyre2006cooperative}.
Depending on system size and simulation settings, common molecular dynamics software packages can reach a few hundreds of ns/day using currently available high-end GPUs~\cite{eastman2017openmm,kutzner2019more}, making these type of calculations unappealing and irrelevant on a pharmaceutical drug discovery timescale.
Other methods compute the free energy of binding by building potential of mean force profiles along a reaction coordinate~\cite{woo2005calculationa,velez-vega2013overcoming,limongelli2013funnel,heinzelmann2017attachpullrelease}, but these methods require prior knowledge of a high-probability binding pathway, which is not easily available, especially in the prospective scenarios typical of the drug development process.

@@

We can rigorously account for fluctuations in other thermodynamic parameters such as changes in volume $V$ when simulating at constant pressure $p$ or changes in number of molecules $N_i$ of species $i$ at constant chemical potential $\mu_i$ (e.g., number of waters or ions) by introducing the \textit{reduced potential}~\cite{shirts2008statisticallya}
\begin{equation}\label{eq:reduced-potential}
u(\vec{q};\vec{\lambda}) \equiv \beta \left[ U(\vec{q};\vec{\lambda}) + p \, V(\vec{q}) + \sum_i \mu_i \, N_i(\vec{q}) + \cdots \right] \, .
\end{equation}
Here, the collection of thermodynamic and alchemical parameters $\{\beta, \vec{\lambda}, p, \mu, \ldots\}$ defines a \emph{thermodynamic state}.
In the context of alchemical calculations, in which the thermodynamic states vary only in their value of $\vec{\lambda}$, these are also referred to as \emph{alchemical states}.
The free energy of mutating $A$ to $B$ in any environment (e.g., binding site, solvent) can then be computed as
\begin{equation}\label{eq:delta-G-env}
    \Delta G_{\mathrm{env}} = - k_BT \ln \frac{Z(\vec{\lambda}_B)}{Z(\vec{\lambda}_A)} = - k_BT \ln \frac{\int_{\Gamma_{\mathrm{env}}} \exp\left( u(\vec{q}; \vec{\lambda}_B) \right) \, d\vec{q}}{\int_{\Gamma_{\mathrm{env}}} \exp\left( u(\vec{q}; \vec{\lambda}_A) \right) \, d\vec{q}} \, .
\end{equation}
While it is generally not feasible to compute the two partition functions $Z(\vec{\lambda})$, several estimators have been devised to robustly estimate the ratio of partition functions in Eq.~\ref{eq:delta-G-env} (see Sec.~\ref{subsec:estimators}) from a set of conformations usually collected with MD simulations from the thermodynamic states defined at $\vec{\lambda}_A$ and $\vec{\lambda}_B$ and intermediates thereof.

>>>>

\newcommand{\versionnumber}{1.0} % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/alchemistry/alchemical-best-practices}} %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Alchemical Free Energy Calculations [Article v \versionnumber]}
\author[1*]{Antonia S. J. S. Mey}
\author[2]{Bryce K. Allen}
\author[3]{Hannah E. Bruce Macdonald}
\author[3*]{John D. Chodera}
\author[9]{David F. Hahn}
\author[1,10]{Maximilian Kuhn}
\author[1]{Julien Michel}
\author[4*]{David L. Mobley}
\author[5]{Levi N. Naden}
\author[6]{Samarjeet Prasad}
\author[2,7]{Andrea Rizzi}
\author[1]{Jenke Scheen}
\author[8*]{Michael R. Shirts}
\author[9]{Gary Tresadern}
\author[2]{Huafeng Xu}
%
\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[2]{Silicon Therapeutics, Boston, MA, USA}
\affil[3]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[4]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, USA}
\affil[5]{Molecular Sciences Software Institute, Blacksburg VA, USA}
\affil[6]{National Institutes of Health, Bethesda, MD, USA}
\affil[7]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[8]{University of Colorado Boulder, Boulder, CO, USA}
\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340,Belgium}
\affil[10]{Cresset, Cambridgeshire, UK}

%
\corr{antonia.mey@ed.ac.uk}{ASJSM}
\corr{john.chodera@choderalab.org}{JDC}
\corr{dmobley@mobleylab.org}{DLM}
\corr{michael.shirts@colorado.edu}{MRS}

\orcid{Antonia S. J. S. Mey}{0000-0001-7512-5252}
\orcid{Bryce Allen}{0000-0002-0804-8127}
\orcid{Hannah E. Bruce Macdonald}{0000-0002-5562-6866}
\orcid{John D. Chodera}{0000-0003-0542-119X}
\orcid{Maximilian Kuhn}{0000-0002-2811-3934}
\orcid{Julien Michel}{0000-0003-0360-1760}
\orcid{David L. Mobley}{0000-0002-1083-5533}
\orcid{Levi N. Naden}{0000-0002-3692-5027}
\orcid{Samarjeet Prasad}{0000-0001-8320-6482}
\orcid{Andrea Rizzi}{0000-0001-7693-2013}
\orcid{Jenke Scheen}{0000-0001-9781-0445}
\orcid{Michael R. Shirts}{0000-0003-3249-1097}
\orcid{Gary Tresadern}{0000-0002-4801-1644}
\orcid{Huafeng Xu}{0000-0001-5447-0452}
\orcid{David F. Hahn}{0000-0003-2830-6880}

\blurb{This LiveCoMS document is maintained online on GitHub at \githubrepository; to provide feedback, suggestions, or help improve it, please visit the GitHub repository and participate via the issue tracker.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PUBLICATION INFORMATION
%%% Fill out these parameters when available
%%% These are used when the "pubversion" option is invoked
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\pubDOI{10.XXXX/YYYYYYY}
\pubvolume{<volume>}
\pubyear{<year>}
\articlenum{<number>}
\datereceived{Month, Day, Year}
\dateaccepted{Month, Day, Year}

@@

In this paper, we review current best practices for several popular application domains of equilibrium alchemical free energy calculations, including relative and absolute small molecule binding free energy calculations to biomolecular targets.
%\todo[inline, color=green!20]{JDC: Should we migrate this document to \url{https://github.com/alchemistry}? ASJSM: Let's migrate with the submitted version, but leave it here for now?}
\end{abstract}
\end{frontmatter}

@@

Subsequent work in the 2000s led to improved implementations of alchemical methods in popular biomolecular simulation packages~\cite{shirts2003extremely,shirts2005solvation,vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace, riniker2011calculation}. 
 This foundational work, combined with the methodological, technological, and hardware improvements of the last 5--10 years, has led to an explosion of interest and direct commercial application of these technologies~\cite{wang2015accurate, fratev2019improved, schindler2020largescale, cournia2017relative, sherborne2016collaborating, kuhn2020assessment}.

@@

As some of the theoretical background can seem daunting, we do, however, provide a guide to the essential theory behind alchemical free energy calculations in Sec.~\ref{sec:theory}.
In the remainder of this paper, we will cover topics that are key to the preparation (Sec.~\ref{sec:prerequisites}), choice and use of correct protocols (Sec.~\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used in the analysis of alchemical calculations (Sec.~\ref{sec:data_analysis}). 
Particular focus will be given to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative). We primarily focus on \emph{equilibrium} free energy calculations in this best practices guide, as best practices for these are more developed, and nonequilibrium techniques may warrant their own guide as such practices evolve.

While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future Best Practices documents focusing on specific issues. 
Below, we provide a non-exhaustive list of topics we have \emph{not} addressed, along with some references to provide starting points on these more advanced topics:
\begin{itemize}
\item covalent inhibition~\cite{lameira2019predicting}
\item free energies of mutation of protein side chains~\cite{gapsys2016accurate, aldeghi2018accurate}
\item nonspecific binding or multiple binding sites~\cite{gill2018binding}
\item approximate and often less accurate endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
\item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
\item forcefield dependence for protein, ligand, ions, co-solvents, and co-factors. A number of different studies have looked at the influence of force fields and it is assumed the user has made an adequate choice for the system under study~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}. 
\item nonequilibrium free energy calculations~\cite{gapsys2020large}
\item Free energy calculations using QM/MM methods~\cite{beierlein2011simple,dybeck2016comparison,cave-ayland2015direct}.
\item Free energy calculations using Machine learning methods~\cite{rufa2020chemical, scheen2020hybrid, cole2020machine}
\end{itemize}

For convenience we have also compiled a list of common acronyms and common symbols used throughout this paper.
\begin{tcolorbox}[title=Acronyms, colback=blue!10!white]
    {\bf CPU} --- Central Processing Unit\\
     {\bf BAR} --- Bennett Acceptance Ratio\\
     {\bf FEP} --- Free Energy Perturbation\\
     {\bf GPCR} --- G-Protein Coupled Receptor\\
     {\bf GPU} --- Graphics Processing Unit\\
     {\bf MBAR} --- Multistate Bennett Acceptance Ratio\\
     {\bf MCSS} --- Maximum Common Substructure\\
     {\bf MD} --- Molecular Dynamics\\
     {\bf RMSE} --- Root Mean Square Error\\
     {\bf MUE} --- Mean Unsigned Error\\
     {\bf SAR} --- Structure-Activity Relationships\\
     {\bf TI} --- Thermodynamic Integration
\end{tcolorbox}

\begin{tcolorbox}[title=List of Symbols, colback=green!10!white]
$L$ and $R$ --- generic names for ligand and receptor\\
$K_b^{\circ}$ --- binding  constant \\
$c^{\circ}$ --- standard state concentration \\
$U$ --- potential energy\\
$u$ --- reduced potential describing a thermodynamic state \\
$\Delta G$ --- Gibbs free energy (isothermal isobaric ensemble), Gibbs function, or free enthalpy, though the most common term Gibbs free energy is used in the text\\
$\Delta A$ --- Helmholtz free energy (canonical ensemble) or Helmholtz function, with Helmholtz free energy used in the text.\\
$\Delta f$ --- reduced (dimensionless) free energy \\
$\Delta \hat{f}$ --- estimate from an estimator for the reduced free energy\\
$\Gamma$ --- configurational space accessible by simulations \\
$\vec{q}$ --- vector of a single configuration, i.e. $x$, $y$, $z$ coordinates of the simulation system\\
$k_B$ --- Boltzmann constant \\
$Z$ --- partition function \\
$p$ --- pressure \\
$\mu$ --- chemical potential (grand canonical ensemble)\\
$T$ --- temperature \\
$\beta \equiv (k_B T)^{-1}$ --- inverse thermal energy \\
$\vec{\lambda}$ --- alchemical progress parameter, which may be multidimensional \\
$g$ --- statistical inefficiency\\
$\mathcal{O}$ --- overlap matrix\\
$C_t$ --- discrete-time-normalized fluctuation auto-correlation function\\
$\tau _{eq}$ --- integrated auto-correlation time\\
$t_0$ --- equilibration time
\end{tcolorbox}

@@

\subsection{Simulating binding events of receptor-drug systems can be computationally expensive}
Suppose you want to compute the binding affinity, or free energy of binding, of a ligand $L$ to a receptor $R$, given by:
\begin{equation}
R+L \leftrightharpoons RL.
\end{equation}
The binding constant ($K_b^{\circ}$) is given by the law of mass action as the ratio of concentrations of product $[RL]$ and reactants $[R]$, $[L]$:
\begin{equation}\label{eq:law-mass-action}
 K_b^{\circ} = c^{\circ}\frac{[RL]}{[L][R]}.
\end{equation}
The standard state concentration $c^\circ$ depends on the reference state, but it is usually set to $1\,\mathrm{mol}/\mathrm{L}$ assuming a constant pressure of $1\,\mathrm{atm}$ (see also Sec.~\ref{sec:standardstate-restraints}), but also holds for dilute solutions if thermodynamic activities can be approximated by concentrations~\cite{gilson1997statisticalthermodynamic}.
Thus, the standard Gibbs free energy of binding $\Delta G_{\mathrm{bind}}$ is given by:
\begin{equation}
    \Delta G_{\mathrm{bind},L} = -k_BT\ln K_b^{\circ},
\end{equation}
where $k_B$ is the Boltzmann constant and $T$ the temperature of the system.
Note that, unless otherwise specified, we use $\Delta G$ throughout the paper to refer to the \textit{standard} free energy of binding (or solvation) (see also Sec.~\ref{sec:standardstate-restraints}), which is often indicated in other works with $\Delta G^{\circ}$ to differentiate it from non-standard free energies.
This is in line with current literature on alchemical calculations, where the standard free energy is normally the only quantity of interest. Furthermore, we will use the term configuration for a single set of position vectors and occasionally use the term conformation to refer to a set of configurations that represent a metastable state.

\paragraph{The free energy of binding can be expressed as a ratio of partition functions}
The law of mass action in Eq.~\ref{eq:law-mass-action} is not directly applicable to typical molecular simulations as they normally include a single receptor/ligand in a small box (i.e., using large concentrations)~\cite{jong2011determining}.
Instead, a natural, though generally very computationally expensive, way to estimate the equilibrium constant is by directly simulating several binding and unbinding events and computing the probability of finding the receptor-ligand system in the bound state, $P(RL)$, or the unbound state, $P(R+L)$.
Assuming the volume change upon binding to be negligible, which is often the case at $1~\mathrm{atm}$ due to the incompressibility of water, then the Gibbs free energy $\Delta G_{\mathrm{bind}, L}$ is approximately equal to the Helmholtz free energy $\Delta A_{\mathrm{bind}, L}$, and we can simulate the system in a box of volume $V$ to obtain~\cite{jong2011determining}
\begin{equation}\label{eq:binding-free-energy-from-bound-unbound-probability-ratio}
    \Delta G_{\mathrm{bind},L} \approx \Delta A_{\mathrm{bind}, L} = - k_B T  \left( \mathrm{ln} \frac{P(RL)}{P(R+L)} + \ln \left( c^\circ N_{\mathrm{Av}} V \right) \right) \, ,
\end{equation}
where $N_{\mathrm{Av}}$ is the Avogadro number, and the last term corrects for the simulated concentration being different than the standard concentration.
Let $\Gamma_{\mathrm{bound}}$ and $\Gamma_{\mathrm{unbound}}$ be the set of receptor-ligand configurations $\vec{q}$ that we consider bound and unbound respectively.
The probability of a configuration $\vec{q}$ is given by the Boltzmann probability density function
\begin{equation}
    P(\vec{q}) = \frac{\exp\left( -\beta U(\vec{q}) \right)}{\int_{\Gamma} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}} \, ,
    \label{eq:conf_probability}
\end{equation}
where $\beta = (k_B T)^{-1}$ is the inverse temperature, $U(\vec{q})$ is the potential energy of configuration $\vec{q}$, and the integration is over the set of all possible configurations accessible in the simulation box volume $\Gamma$, with $\Gamma_{\mathrm{bound}},\Gamma_{\mathrm{unbound}} \subset \Gamma$.
If the simulation is long enough, we expect the fraction of configurations $\vec{q}$ found in the bound state to converge to
\begin{equation}
    P(RL) = \int_{\Gamma_{\mathrm{bound}}} P(\vec{q}) \, d\vec{q} = \frac{\int_{\Gamma_{\mathrm{bound}}} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}}{\int_{\Gamma} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}} \, .
\end{equation}
After similar considerations for $P(R+L)$, we find that the ratio of visited bound and unbound conformations, in the limit of long simulations, should converge to
\begin{equation}\label{eq:bound-unbound-probability-ratio}
    \frac{P(RL)}{P(R+L)} = \frac{\int_{\Gamma_{\mathrm{bound}}} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}}{\int_{\Gamma_{\mathrm{unbound}}} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}} = \frac{Z(RL)}{Z(R+L)} \, ,
\end{equation}
where we have defined the \textit{configurational integral} or \textit{configurational partition function} as $Z(\mathrm{state}) \equiv \int_{\Gamma_{\mathrm{state}}} \exp\left(-\beta U(\vec{q})\right) \, d\vec{q}$.

\paragraph{Simulating binding events is computationally expensive}
While simulating binding events has been used to estimate binding affinities~\cite{jong2011determining,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo2016adaptive,votapka2017seekr,doerr2014onthefly,plattner2015protein,dixon2018predicting}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the microsecond timescale even for millimolar binders~\cite{pan2017quantitative} and reaches the microsecond to second timescale for a typical drug~\cite{basavapathruni2012conformational,hyre2006cooperative}.
Depending on system size and simulation settings, common molecular dynamics software packages can reach a few hundreds of ns/day using currently available high-end GPUs~\cite{eastman2017openmm,kutzner2019more}, making these type of calculations unappealing and irrelevant on a pharmaceutical drug discovery timescale.
Other methods compute the free energy of binding by building potential of mean force profiles along a reaction coordinate~\cite{woo2005calculationa,velez-vega2013overcoming,limongelli2013funnel,heinzelmann2017attachpullrelease}, but these methods require prior knowledge of a high-probability binding pathway, which is not easily available, especially in the prospective scenarios typical of the drug development process.

@@

We can rigorously account for fluctuations in other thermodynamic parameters such as changes in volume $V$ when simulating at constant pressure $p$ or changes in number of molecules $N_i$ of species $i$ at constant chemical potential $\mu_i$ (e.g., number of waters or ions) by introducing the \textit{reduced potential}~\cite{shirts2008statisticallya}
\begin{equation}\label{eq:reduced-potential}
u(\vec{q};\vec{\lambda}) \equiv \beta \left[ U(\vec{q};\vec{\lambda}) + p \, V(\vec{q}) + \sum_i \mu_i \, N_i(\vec{q}) + \cdots \right] \, .
\end{equation}
Here, the collection of thermodynamic and alchemical parameters $\{\beta, \vec{\lambda}, p, \mu, \ldots\}$ defines a \emph{thermodynamic state}.
In the context of alchemical calculations, in which the thermodynamic states vary only in their value of $\vec{\lambda}$, these are also referred to as \emph{alchemical states}.
The free energy of mutating $A$ to $B$ in any environment (e.g., binding site, solvent) can then be computed as
\begin{equation}\label{eq:delta-G-env}
    \Delta G_{\mathrm{env}} = - k_BT \ln \frac{Z(\vec{\lambda}_B)}{Z(\vec{\lambda}_A)} = - k_BT \ln \frac{\int_{\Gamma_{\mathrm{env}}} \exp\left( u(\vec{q}; \vec{\lambda}_B) \right) \, d\vec{q}}{\int_{\Gamma_{\mathrm{env}}} \exp\left( u(\vec{q}; \vec{\lambda}_A) \right) \, d\vec{q}} \, .
\end{equation}
While it is generally not feasible to compute the two partition functions $Z(\vec{\lambda})$, several estimators have been devised to robustly estimate the ratio of partition functions in Eq.~\ref{eq:delta-G-env} (see Sec.~\ref{subsec:estimators}) from a set of configurations usually collected with MD simulations from the thermodynamic states defined at $\vec{\lambda}_A$ and $\vec{\lambda}_B$ and intermediates thereof.
